---
title: "American Election 2020 prediction using Logistic Regression"
author: "Danny Xu, Nayoung Kim, Steven Tran"
date: "November 2, 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(tidyverse)
library(survey)

# Loading in the cleaned survey Data
survey_data <- read_csv("survey_data.csv") %>% drop_na

# Loading in the cleaned census Data
census_data <- read_csv("census_data.csv")

# Ordering some variables as factors in order to perform logistic regression
survey_data$age_bin <- factor(survey_data$age_bin, levels=
                          c("18 to 24",
                            "25 to 34",
                            "35 to 44",
                            "45 to 54",
                            "55 to 64",
                            "65 to 74",
                            "75 or older"
                            ))
survey_data$education <- factor(survey_data$education, levels=
                          c("3rd Grade or less",
                            "Middle School - Grades 4 - 8",
                            "Completed some high school",
                            "High school graduate",
                            "Associate Degree",
                            "Completed some college, but no degree",
                            "College Degree (such as B.A., B.S.)",
                            "Masters degree",
                            "Doctorate degree"
                            ))
survey_data$household_income <- factor(survey_data$household_income, levels=
                          c("Less than $14,999",
                            "$15,000 to $19,999",
                            "$20,000 to $24,999",
                            "$25,000 to $29,999",
                            "$30,000 to $34,999",
                            "$35,000 to $39,999",
                            "$40,000 to $44,999",
                            "$45,000 to $49,999",
                            "$50,000 to $54,999",
                            "$55,000 to $59,999",
                            "$60,000 to $64,999",
                            "$65,000 to $69,999",
                            "$70,000 to $74,999",
                            "$75,000 to $79,999",
                            "$80,000 to $84,999",
                            "$85,000 to $89,999",
                            "$90,000 to $94,999",
                            "$95,000 to $99,999",
                            "$100,000 to $124,999",
                            "$125,000 to $149,999",
                            "$150,000 to $174,999",
                            "$175,000 to $199,999",
                            "$200,000 to $249,999",
                            "$250,000 and above"
                            ))
survey_data$gender <- factor(survey_data$gender, levels=
                          c("Female",
                            "Male"
                            ))
survey_data$foreign_born <- factor(survey_data$foreign_born, levels=
                          c("Another country",
                            "The United States"
                            ))
```

<!-- # American Election 2020 prediction using Logistic Regression

## Name(s) of Author(s)
## November 2, 2020 -->

# Introduction

Many Americans and people across the world alike have made strong predictions about the popular vote outcome of the 2020 American federal election. We are interested in predicting the outcome using a logistic regression model fitted to survey data, and employing a post-stratification technique using US census data. In this report, we will discuss the model specifics, how the population data was used to weigh the strata and make them more representative, and finally, the predicted winner as determined by our methods.

## Data

Two datasets were used in this report. For the survey data, we selected a subset of the Nationscape Data Set published by the Democracy Fund and UCLA Nationscape groups for the most recent survey that was conducted on June 25th 2020 (**citation needed here**). Among other variables discussed below, respondents were asked whether they would vote for Donald Trump or Joe Biden on that date. The census data is provided by IPUMS as the 'American Community Survey 2014-2018 5-Year Sample'; we used, along with the default variables which included strata weights, the following attributes: `HHINCOME, US2018C_ST, SEX, RACE, BPL, US2018C_SCHL, AGE`. These variables were selected because equivalent questions were asked in the survey, and they describe the demographics of the respondents well. The data was cleaned into matching observation values in order to obtain a correspondence between the two datasets for later use when performing the post-stratification.

# Model

## Model Specifics
We will be using a logistic regression model to model the proportion of voters who will vote for Donald Trump. A logistic model is suitable because the event of voting for Donald Trump is a Bernoulli random variable which means that it can only take on two possible values. We will be using birthplace, gender, race/ethnicity, household income, education, state, and age to model the probability of voting for Donald Trump. The variables are ordered in the following way:

- Ordinal variables are:
  - `income`:
    - baseline: "`r levels(survey_data$household_income)[1]`"
    - order: (`r levels(survey_data$household_income) %>% sapply(function(x) toString(dQuote(x, F)))`)
  - `education`:
    - baseline: "`r levels(survey_data$education)[1]`""
    - order: (`r levels(survey_data$education) %>% sapply(function(x) toString(dQuote(x, F)))`)
  - `age_bin`:
    - baseline: "`r levels(survey_data$age_bin)[1]`"
    - order: (`r levels(survey_data$age_bin) %>% sapply(function(x) toString(dQuote(x, F)))`)
  - `gender`:
    - baseline: "`r levels(survey_data$gender)[1]`"
    - order: (`r levels(survey_data$gender) %>% sapply(function(x) toString(dQuote(x, F)))`)
  - `foreign_born`:
    - baseline: "`r levels(survey_data$foreign_born)[1]`"
    - order: (`r levels(survey_data$foreign_born) %>% sapply(function(x) toString(dQuote(x, F)))`)
- Nominal variables have no order and are: `race` (`r survey_data$race_ethnicity %>% unique`), and `state` (50 states in the US).

Thus, the logistic regression model we are using is:

\begin{align*}
    \ln \left(\frac{P_T}{1-P_T}\right) &= \beta_0+\beta_{\text{US}}x_{\text{US}} +\beta_{\text{male}}x_{\text{male}} \\
                                       &+\sum_{\substack{
            i\in \text{race}, \\
            i\neq \text{`American Indian'}
            }}{\beta_ix_i} + \sum_{\substack{
            i\in \text{income}, \\
            i\neq \text{`Less than \$14,999'}
            }}{\beta_ix_i} + \sum_{\substack{
            i\in \text{education}, \\
            i\neq \text{`3rd grade or less'}
            }}{\beta_ix_i} \\
                                       &+ \sum_{\substack{
            i\in \text{state}, \\
            i\neq \text{`AK'}
            }}{\beta_ix_i} + \sum_{\substack{
            i\in \text{age}, \\
            i\neq \text{`18 to 24'}
    }}{\beta_ix_i}
\end{align*}

Where $P_T$ represents the proportion of voters who will vote for Donald Trump. Similarly, the intercept term $\beta_0$ represents the baseline log-odds of voting for Donald Trump; that is, the voter is born outside the US, is female and American Indian, has a household income less than \$14,999, below third grade education, living in Arkansas, and is between 18 and 24 years old. With this dummy variable coding, changing the 'level' of a variable by changing the value of one $x_i$ from 1 to 0 would affect the log-odds by $\beta_i$, the regression estimate for a certain variable. In other words, they represent the average change in log-odds when going from the default value of the variable to the $i$th value of the variable.

The default values for all the variables were listed previously in the explanation of $\beta_0$. As a specific example, $\beta_{white}$ represents the average change in log-odds of voting for Donald Trump when going from American Indian observations to White observations. In this regression model, we used sampling weights which give the proportion of observations in the population that the observation represents. These sampling weights are automatically normalized according to the sample size as they have been provided for us in the raw survey dataset.

We will also use a similar logistic regression model to model the proportion of voters who will vote for Joe Biden. The explanatory variables are the exact same as the previous model, but the log-odds becomes
$$ \ln\left(\frac{P_B}{1-P_B}\right) $$
Where $P_B$ represents the proportion of voters who will vote for Joe Biden. Interpretation of the parameters for the Joe Biden model are analogous to those for the model used to predict Donald Trump's vote proportions.

```{r, include=F}
# Creating the Model
model.design <- svydesign(ids=~1, data = survey_data, weights = ~weight)

trump_model <- svyglm(vote_trump ~ foreign_born + gender + race_ethnicity + household_income + education + state + age_bin,
                      family = quasibinomial,
                      design = model.design)

# Model Results (to Report in Results section)
summary(trump_model)

biden_model <- svyglm(vote_biden ~ foreign_born + gender + race_ethnicity + household_income + education + state + age_bin,
                      family = quasibinomial,
                      design = model.design)

summary(biden_model)
```

## Post-Stratification

In order to estimate the proportion of voters who will vote for Donald Trump and Joe Biden we will need to perform a post-stratification analysis. Here we create cells based off each of the explanatory variables in our model: birthplace, gender, race/ethnicity, household income, education, state, and age. We chose variables that we thought affected political leaning and were also in both both datasets. Using the models described in the previous sub-section we will estimate the proportion of voters who will vote for Donald Trump and Joe Biden in each cell. We will then weight each proportion estimate (within each cell) by the respective population size of that cell and sum those values and divide that by the entire population size. We will perform this twice, once to obtain the post-stratified proportion of voters who will vote for Donald Trump, and another to obtain the post-stratified proportion of voters who will vote for Joe Biden. In doing this post-stratification, we obtain a more accurate estimate of the proportion of voters who will vote for Donald Trump and Joe Biden. This is achieved by using the much larger census dataset to reduce the bias our smaller survey dataset may have.

```{r, include=F}

# Perform the post-stratification calculations
census_data$logodds_estimate <- trump_model %>%
  predict(newdata = census_data)

census_data$estimate <- exp(census_data$logodds_estimate)/(1+exp(census_data$logodds_estimate))

trump_prop <- census_data %>%
  mutate(alp_predict_prop = estimate*n) %>%
  summarise(alp_predict = sum(alp_predict_prop)/sum(n))

trump_prop <- as.numeric(trump_prop)


census_data$logodds_estimate <- biden_model %>%
  predict(newdata = census_data)

census_data$estimate <- exp(census_data$logodds_estimate)/(1+exp(census_data$logodds_estimate))

biden_prop <- census_data %>%
  mutate(alp_predict_prop = estimate*n) %>%
  summarise(alp_predict = sum(alp_predict_prop)/sum(n))

biden_prop <- as.numeric(biden_prop)

```


# Results

Here you will include all results. This includes descriptive statistics, graphs, figures, tables, and model results. Please ensure that everything is well formatted and in a report style. You must also provide an explanation of the results in this section.

Please ensure that everything is well labelled. So if you have multiple histograms and plots, calling them Figure 1, 2, 3, etc. and referencing them as Figure 1, Figure 2, etc. in your report will be expected. The reader should not get lost in a sea of information. Make sure to have the results be clean, well formatted and digestible.

We calculated that the proportion of voters who would vote for Donald Trump is 0.374. This was calculated using the logistic regression model we developed for predicting the log-odds of voters who would vote for Donald Trump, which took into account the voter's birthplace, gender, race/ethnicity, household income, education, state, and age. Similarly, we calculated that the proportion of voters who would vote for Joe Biden is 0.408.

# Discussion

Here you will summarize the previous sections and discuss conclusions drawn from the results. Make sure to elaborate and connect your analysis to the goal of the study.

Summary:

Conclusions:
Using the estimated proportion of voters who would vote for Donald Trump and comparing against the proportion of voters who would vote for Joe Biden shows that the latter is larger. Using this, we predict that Joe Biden will win the presidential election.

## Weaknesses

Here we discuss weaknesses of the study, data, analysis, etc. You can also discuss areas for improvement.

Comparing education across datasets was a challenge. Vocational schooling and graduate without degree were possible values in the survey dataset, but had no corresponding values in the census dataset. We decided to filter out all observations with these values, to ensure we could post-stratify. However in doing so, we may have introduced more bias to our already small dataset. We reduced the dataset size from approximately 6500 observations to approximately 6000 observations, which decreases the accuracy of our estimates in our model. In addition, removing specific values from our dataset may skew towards a certain political candidate. Professional degree beyond bachelor's degree was a category in education for the census data that had no survey data equivalent observations in this category were also removed. Similarly, we encountered problems in the race variable across datasets. The census dataset had categories for multiple major races, but the survey dataset did not. We decided to remove all observations in the census dataset that were in these categories. This could introduce bias since it could be that people who are in multiple major race categories could prefer to vote Biden over Trump. In addition, we removed observations in the census data that did not report their income, due to the survey data having no such category. The exact meanings of the variables across datasets was also not the same. Survey data only had annual household income before taxes, while census data also incorporated losses. This can possibly skew the data because losses would shift incomes down, so overall more people would have lower annual household incomes (relative to the model), which might overestimate Biden votes over Trump votes.

## Next Steps

Here you discuss subsequent work to be done after this report. This can include next steps in terms of statistical analysis (perhaps there is a more efficient algorithm available, or perhaps there is a caveat in the data that would allow for some new technique). Future steps should also be specified in terms of the study setting (eg. including a follow-up survey on something, or a subsequent study that would complement the conclusions of your report).

After the election results are determined and available, one can determine the accuracy of the model in this paper by examining our estimates against the election results. Particularly, examining the ratio of the proportion of Trump voters and the proportion of Biden voters of both the model and the election results will show us how accurate the model was. We can also break down the votes by state and determine if it was a good variable to include in the logistic regression model or would something else like districts be better.

# References

1. Tausanovitch, Chris and Lynn Vavreck. 2020. Democracy Fund + UCLA Nationscape, October 10-17, 2019 (version 20200814). Retrieved from https://www.voterstudygroup.org/publication/nationscape-data-set.

2. Steven Ruggles, Sarah Flood, Ronald Goeken, Josiah Grover, Erin Meyer, Jose Pacas and Matthew Sobek. IPUMS USA: Version 10.0 AMERICAN COMMUNITY SURVEY 2014-2018 5-YEAR SAMPLE. Minneapolis, MN: IPUMS, 2020. https://doi.org/10.18128/D010.V10.0

Make sure to cite R and every package.
